A Domain-Specific Optimizing Compiler\\for Solving Partial Differential Equations

### 2014-02-14 paper/bare_conf.tex
Write out a first version of the introduction

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Domain-Driven Code Transformation\\for Solving Partial Differential Equations}

@@

\section{Introduction}
In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, programs are required that model phenomena. A feature common to these scientific computations is that they involve the resolution of partial differential equations (PDEs). Numerical techniques, like Finite Volume Method (FVM) and Finite Element Method (FEM), are widely employed in real-world applications to approximate solutions of PDEs. Unstructured meshes are often used to discretize the PDE domain, since their geometric flexibility allows such PDE solvers to be extrimely effective. However, they are also the main responsible for performance degradation due to the introduction of indirect accesses (e.g. \texttt{A[B[i]]}) to access mesh components. A typical problem, therefore, is that implementing a scientific application requires a great effort, other than domain knowledge, and even more time is needed to optimize their execution time.

An unstructured mesh represents a discretization of the equation domain into two- or three-dimensional cells (or ``elements''), which define the smallest area within which the solution is sought. As the number of elements can be of the order of millions, a major issue is the time required to perform the numerical method, which can be of the order of hours or days. OP2~\cite{pyop2ics} and Lizst~\cite{lizst} are Domain Specific Languages that target mesh-based computations by allowing programmers to achieve highly-efficient parallel execution. The successful porting of Hydra, a CFD industrial application devised by Rolls Royce for turbomachinery design (based on FVM, roughly 50000 lines of code and a mesh size that can be over 100 millions edges), to OP2, demonstrates the effectiveness of this framework for implementing PDEs solvers~\cite{OP2-hydra}. Both OP2 and Lizst adopt a kernel-oriented programming model: the computation semantics is expressed through self-contained functions (or ``kernels''), each function applied in parallel to all items in a specific set of mesh components (e.g. edges, vertices, elements), with an implicit synchronization between the application of two different functions. Together with the indirections problem, having systematic optimizations for the kernels is the major concern in unstructured mesh applications.

In this paper, we describe a model-driven optimization strategy for a class of kernels used in Finite Element computations. The working set of these kernels is usually small enough to fit the L1 cache, and definitely fits the L2 cache when high-order methods are employed to improve the accuracy of the solution. However, the latter case is not considered in our work. This implies that main actors of our study will be register allocation and instruction-level parallelism, especially in the form of SIMD vectorisation.

@@

The compiler cost model. Used to find out tile size and split size.

>>>>

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{A Domain-Specific Optimizing Compiler\\for Solving Partial Differential Equations}

@@

\section{Introduction}
In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, programs model phenomena by means of partial differential equations (PDEs). Numerical techniques, like Finite Volume Method (FVM) and Finite Element Method (FEM), are widely employed in real-world applications to approximate solutions of PDEs. Unstructured meshes are often used to discretize the domain of a PDE, since their geometric flexibility allows solvers to be extrimely effective. However, they are also one of the main responsible for performance degradation, due to the introduction of indirect accesses (e.g. \texttt{A[B[i]]}) to access mesh components. A typical problem, therefore, is that implementing a scientific application requires a great effort, other than domain knowledge, and even more time is needed to optimize their execution time.

An unstructured mesh discretizes the domain of the equation into two- or three-dimensional cells (or ``elements''), within which the solution is sought. As the number of elements can be of the order of millions, a major issue is the time required to perform the numerical method, which can be hours or days. To address this issue, domain-specific languages (DSLs), which ease the programming burden and enable highly-efficient parallel execution, have been developed. The successful porting of Hydra, a CFD industrial application devised by Rolls Royce for turbomachinery design (based on FVM, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to the OP2 DSL~\cite{pyop2ics}, demonstrates the effectiveness of the DSL approach to implementing PDEs solvers~\cite{OP2-hydra}.

OP2, as well as Lizst~\cite{lizst}, adopt a kernel-oriented programming model, in which the computation semantics is expressed through self-contained functions (or ``kernels''). A kernel is applied to all elements in a set of mesh components (e.g. edges, vertices, elements), with an implicit synchronization between the application of two consecutive kernels. On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread. Together with the indirections problem, kernel optimization is the major concern in unstructured mesh applications. In this paper, we tackle this problem by proposing a domain-driven optimization strategy for a class of kernels used in FEM.

We focus on Finite Element Local Assembly (``assembly'', in the following), a fundamental step of FEM that covers an important fraction of the overall computation run-time, typically in the range 30$\%$-60$\%$. During the assembly phase, the solution of the PDE by executing a suitable kernel over all elements in the discretized domain. A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when (very) high-order methods are employed to improve the accuracy of the solution. However, we do not consider the latter case. An assembly kernel is characterized by the presence of an affine, usually non-perfect loop nest, where individual loops are rather small (the trip count rarely exceeds 100, with a minimum value of 3, depending on the order of the method). With such small kernels, our work focuses on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorisation.

Our study is conducted in the context of Firedrake, a multilayered system for the automated solution of PDEs through FEM. PDEs are written in Firedrake using a domain-specific language called Unified Form Language (UFL). The high-level UFL notation is used by the Fenics Form Compiler (FFC) to produce assembly kernels. At the core of Firedrake~\cite{firedrake} there is PyOP2~\cite{pyop2ics} - a python implementation of the OP2 abstraction - which triggers compilation of kernels using an available vendor compiler, and eventually manages parallel execution over the unstructured mesh. We exploit the domain knowledge and the structure inherent to FFC's assembly kernels to develop an optimizing compiler for all problems that are expressible in Firedrake.

The compiler we developed is integrated with PyOP2, and already used in the release of Firedrake~\cite{firedrake-code}. This allows us to evaluate the proposed code transformations in a range of real-world problems, employing different low-order methods (from $p=1$ to $p=4$; the higher, the more expensive a kernel is, but better is the accuracy), and varying the topology of the mesh (2D, 3D, 3D hybrid structured-unstructured, which in turn affect the kernel cost). Performance results show that compiler-generated code for non-trivial assembly kernels, if not adequately transformed, is sub-optimal. A cost-model-driven sequence of source-to-source code transformations, aimed at improving register allocation and register data locality, can result in performance improvements up to 33$\%$ over ``softly-optimized'' code (i.e. where only basic transformations are performed, such as auto-vectorizable loop-invariant code motion, padding and data alignment), and up to 77$\%$ over the original FFC kernels. The contribution of this paper is therefore twofold:
\begin{itemize}
\item An optimisation strategy for a class of kernels widely-used in scientific applications, namely Finite Element Local Assembly kernels. Our approach exploits domain knowledge to go beyond the limits of both vendor (e.g. $icc$) and research (e.g. those based on the polyhedral model) compilers.
\item Automation of such code optimizations in Firedrake, a framework for the resoloution of PDEs through FEMs. This enables performing an in-depth performance evaluation of the proposed optimization strategy in real-world simulations.
\end{itemize}

The paper is organized as follows. ...

@@

The compiler cost model. Used to find out suitable unroll-and-jam factor and split size.
